WASP-4\lowercase{b} Arrived Early for the TESS Mission

### 2019-03-27 paper/ms.tex
minor text edits

\NewPageAfterKeywords
\listofchanges

@@

WASP-4 was observed by TESS with Camera 2 from August 23 to September
20, 2018, within the second ``sector'' of science operations.  The
star is designated as TIC 402026209 in the TESS Input Catalog
\citep{stassun_TIC_2018}.  The pixel data for an $11\times11$ array
surrounding WASP-4 were averaged into 2-minute stacks by the onboard
computer.  The data were downlinked via the Deep Space
Network\footnote{\url{deepspace.jpl.nasa.gov}}, and the
spacecraft timestamps were calibrated against the ground-station
clocks.  The spacecraft clock times were then transformed by the
Payload Operations Center into the {\it Temps Dynamique Barycentrique}
(TDB) reference system.  The images were then reduced to lightcurves
by the Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  During this processing, the SPOC used
the known spacecraft trajectory to compute the barycentric time
corrections on a target by target basis, and expressed the timestamps
as Barycentric TESS Julian Dates (BTJD), which is simply the
\replaced{b}{B}arycentric Julian \replaced{d}{D}ate minus
2{,}457{,}000.  Lightcurves that were flagged \deleted{as Threshold
Crossing Events} by the SPOC pipeline \added{as crossing a transit
detection threshold} were then vetted and released by the MIT TESS
Science Office to the Mikulski Archive for Space Telescopes on
November 29, 2018~\citep{ricker_tess_alerts_2018}.

@@

\added{We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions. The prior for
the quadratic model allowed the period derivative to have any sign.
We considered two possible priors for the precession model: the first
is a wide prior that allows non-physical values of the planetary
Love number (Equation~\ref{eq:love_number}). The second requires
that the planetary Love number is less than that of a sphere with a
homogeneous density profile.  }

Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
\replaced{decreasing-period}{quadratic} model has $\chi^2 = 62.6$ and
60 degrees of freedom.  The best-fitting precession model \added{under
the wide prior} has \replaced{$\chi^2 = 64.3$}{$\chi^2 = 64.0$} and 58
degrees of freedom.\added{ Under the physical prior, the fit is slightly
worse, with $\chi^2=64.8$}. The precession and quadratic models both
provide much better fits than the constant-period model.  The
difference in $\chi^2$ between the linear and quadratic models
corresponds to $p \approx 10^{-26}$.

The \replaced{decreasing-period}{quadratic} model provides a slightly
better fit to the data than the precession model.  It is favored by
\replaced{$\Delta \chi^2 = 1.7$}{$\Delta \chi^2 = 1.4$}, and has two
fewer free parameters.  A useful heuristic for model comparison is the
Bayesian Information Criterion,
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
prec} - {\rm BIC}_{\rm quad} = 10$\added{, assuming a wide prior for
the precession model}. This corresponds to a Bayes factor of
$1.1\times 10^{4}$.  Likewise, the Akaike Information Criterion favors
the constant-period-derivative model by $\Delta {\rm AIC} = 5.8$.
Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

@@

\explain{The result given in Equation~\ref{eq:love_number} has been
corrected from the submitted version. The previously stated result, k2
= $1.5^{+2.1}_{-1.1}$ had a whopping three errors. The first was a
typo that led to printing of 84th and 16th percentiles, instead of the
difference of these percentiles from the median. The second was an
insufficiently precise specification of the assumed priors, which are
now described.  The third was ignoring the uncertainty in the planet
radius and semimajor axis, which enter at the fifth power and are now
included.  }

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 27 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{align}
  k_{2,{\rm p}} &= 1.59^{+0.70}_{-0.47}
    \quad{\rm (wide\ prior),} \\
  k_{2,{\rm p}} &=  1.20^{+0.20}_{-0.26}
    \quad(k_{2,{\rm p}} \sim \mathcal{U}[0.015, 1.5])
  \label{eq:love_number}
\end{align}
where the uncertainties in the semimajor axis and planet radius have been
propagated through our Markov chains, and $\mathcal{U}$ denotes the uniform 
distribution.  For comparison, the Love number
of Jupiter is about 0.55 \citep{wahl_tidal_2016,ni_empirical_2018},
and a uniform density sphere has $k_2 = 1.5$. The uncertainty in
$k_{2,{\rm p}}$ for WASP-4b is large because the eccentricity,
reference time, and ${\rm d}\omega/{\rm d}E$ have strongly correlated
errors, and the \added{four available}\deleted{measured} occultation
times only weakly constrain these parameters
(Figure~\ref{fig:future}).

\added{ The results under both priors are larger than the Love number
of Jupiter, at weak statistical signifiance.  Imposing the
physically-motivated prior truncates the allowed values of $k_{2,{\rm
p}}$, and drives the precession period towards larger values.
Different priors also change the predicted evolution of the
orbit.  While the precession model matches the orbital decay model for
the next two decades in transits (Figure~\ref{fig:future}), in
occultations the precession model can deviate from the orbital decay
model, particularly if one imposes stricter priors on the planetary
Love number. This suggests that promptly obtaining occultation time
measurements for the system could help rule out the apsidal precession
model.  For the time being, the data are insufficient to make a
stronger ruling.  }

The main problem with the apsidal precession hypothesis \added{under
either of the priors} is to explain why the eccentricity would be as
large as $\sim$$10^{-3}$ despite rapid tidal circularization.  For
WASP-4, Equation~\ref{eq:de_dt} gives $\tau_e = 0.29 (Q_{\rm p}'/10^5)
\, {\rm Myr}$.  The star is several billion years old, so unless the
planet arrived very recently, any initial eccentricity should have
been lowered well below $\sim$$10^{-3}$.  The top panel of
Figure~\ref{fig:context} compares the expected eccentricity damping
time of WASP-4b with that of other transiting giant planets.  WASP-4b
has one of the shortest known eccentricity damping times.

>>>>

%\NewPageAfterKeywords
\listofchanges

@@

WASP-4 was observed by TESS with Camera 2 from August 23 to September
20, 2018, within the second ``sector'' of science operations.  The
star is designated as TIC 402026209 in the TESS Input Catalog
\citep{stassun_TIC_2018}.  The pixel data for an $11\times11$ array
surrounding WASP-4 were averaged into 2-minute stacks by the onboard
computer.  The data were downlinked via the Deep Space
Network\footnote{\url{deepspace.jpl.nasa.gov}}, and the
spacecraft timestamps were calibrated against the ground-station
clocks.  The spacecraft clock times were then transformed by the
Payload Operations Center into the {\it Temps Dynamique Barycentrique}
(TDB) reference system.  The images were then reduced to lightcurves
by the Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  During this processing, the SPOC used
the known spacecraft trajectory to compute the barycentric time
corrections on a target by target basis, and expressed the timestamps
as Barycentric TESS Julian Dates (BTJD), which is simply the
\replaced{b}{B}arycentric Julian \replaced{d}{D}ate minus
2{,}457{,}000.  Lightcurves that were flagged\deleted{ as Threshold
Crossing Events} by the SPOC pipeline \added{as crossing a transit
detection threshold} were then vetted and released by the MIT TESS
Science Office to the Mikulski Archive for Space Telescopes on
November 29, 2018~\citep{ricker_tess_alerts_2018}.

@@

\added{We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions. The prior for
the quadratic model allowed the period derivative to have any sign.
We considered two possible priors for the precession model: the first
is a wide prior that allows non-physical values of the planetary
Love number (Equation~\ref{eq:love_number}). The second prior requires
that the planetary Love number is less than that of a sphere of
constant density.  }

Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
\replaced{decreasing-period}{quadratic} model has $\chi^2 = 62.6$ and
60 degrees of freedom.  The best-fitting precession model \added{under
the wide prior} has $\chi^2 = 64.3$ and 58
degrees of freedom.\added{ Under the physical prior, the best-fit is
slightly worse, with $\chi^2=64.8$}. The precession and quadratic
models both provide much better fits than the constant-period model.
The difference in $\chi^2$ between the linear and quadratic models
corresponds to $p \approx 10^{-26}$.

The \replaced{decreasing-period}{quadratic} model provides a
marginally better fit to the data than the precession model.  It is
favored by $\Delta \chi^2 = 1.7$, and has two fewer free parameters.
A useful heuristic for model comparison is the Bayesian Information
Criterion,
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
prec} - {\rm BIC}_{\rm quad} = 10$\added{, assuming a wide prior for
the precession model}. This corresponds to a Bayes factor of
$1.1\times 10^{4}$.  Likewise, the Akaike Information Criterion favors
the constant-period-derivative model by $\Delta {\rm AIC} = 5.8$.
Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

@@

\explain{The result given in Equation~\ref{eq:love_number} has been
corrected from the submitted version. The previously stated result, k2
= $1.5^{+2.1}_{-1.1}$ had a
typo that led to printing of 84th and 16th percentiles, instead of the
difference of these percentiles from the median. 
The uncertainties also previously did not propagate uncertainty in the
planet radius and semimajor axis, which enter at the fifth power and
are now included. 
Finally, we now also specify the assumed priors, as they add a
nuance in the interpretation.   }

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 27 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{align}
  k_{2,{\rm p}} &= 1.59^{+0.70}_{-0.47}
    \quad{\rm (wide\ prior),} \label{eq:love_number} \\
  k_{2,{\rm p}} &=  1.20^{+0.20}_{-0.26}
    \quad(k_{2,{\rm p}} \sim \mathcal{U}[0.015, 1.5])
\end{align}
where the uncertainties in the semimajor axis and planet radius have been
propagated through our Markov chains, and $\mathcal{U}$ denotes the uniform 
distribution.  For comparison, the Love number
of Jupiter is about 0.55 \citep{wahl_tidal_2016,ni_empirical_2018},
and a uniform density sphere has $k_2 = 1.5$. The uncertainty in
$k_{2,{\rm p}}$ for WASP-4b is large because the eccentricity,
reference time, and ${\rm d}\omega/{\rm d}E$ have strongly correlated
errors, and the \added{four available}\deleted{measured} occultation
times only weakly constrain these parameters
(Figure~\ref{fig:future}).

\added{ The results under both priors are larger than the Love number
of Jupiter, at weak statistical signifiance.  Imposing the
physically-motivated prior truncates the allowed values of $k_{2,{\rm
p}}$, and drives the precession period towards larger values.
Different priors therefore change the predicted evolution of the
orbit.  While the precession model matches the orbital decay model for
the next two decades in transits (Figure~\ref{fig:future}), in
occultations the precession model can deviate from the orbital decay
model, particularly if one requires the planetary Love number to be
smaller. This suggests that promptly obtaining occultation time
measurements for the system could help rule out the apsidal precession
model.  For the time being, the data are insufficient to make a
stronger judgement.  }

\added{Regardless of the detailed model assumptions, the}
\deleted{The} main \added{physical }problem with the apsidal precession
hypothesis is to explain why the
eccentricity would be as large as $\sim$$10^{-3}$ despite rapid tidal
circularization.  For WASP-4, Equation~\ref{eq:de_dt} gives $\tau_e =
0.29 (Q_{\rm p}'/10^5) \, {\rm Myr}$.  The star is several billion
years old, so unless the planet arrived very recently, any initial
eccentricity should have been lowered well below $\sim$$10^{-3}$.  The
top panel of Figure~\ref{fig:context} compares the expected
eccentricity damping time of WASP-4b with that of other transiting
giant planets.  WASP-4b has one of the shortest known eccentricity
damping times.
